\chapter{Results and Evaluation}
\label{ch:results}

This chapter presents the complete experimental results of this thesis, progressing from the controlled \gls{ann}--\gls{snn} comparison through spike encoding analysis, neuromorphic hardware deployment, energy measurement, transfer learning, and advanced properties including adversarial robustness and continual learning. Every reported mean and standard deviation is computed across the five predefined ESC-50 cross-validation folds unless otherwise stated. All software experiments were conducted on NVIDIA A100 SXM4 80\,GB GPUs on the University of Manchester CSF3 cluster; all hardware experiments used the SpiNN-5 board at \texttt{spinnaker.cs.man.ac.uk}.

% ==================================================================
\section{ANN Baseline}
\label{sec:ann-baseline}
% ==================================================================

The ConvANN baseline establishes the upper bound for what a non-spiking network of identical architecture and training budget can achieve on ESC-50 without external pretraining. Table~\ref{tab:ann-baseline} reports the per-fold results.

\begin{table}[t]
\centering
\caption{ConvANN baseline accuracy (\%) on ESC-50 with five-fold cross-validation. The model has $\sim$622K parameters and is trained from scratch on mel spectrograms.}
\label{tab:ann-baseline}
\small
\begin{tabular}{@{}lccccccc@{}}
\toprule
\textbf{Model} & \textbf{Fold 1} & \textbf{Fold 2} & \textbf{Fold 3} & \textbf{Fold 4} & \textbf{Fold 5} & \textbf{Mean} & \textbf{Std} \\
\midrule
ConvANN & 66.25 & 64.00 & 68.75 & 60.00 & 60.25 & 63.85 & 3.07 \\
\bottomrule
\end{tabular}
\end{table}

The mean accuracy of 63.85\% $\pm$ 3.07\% is substantially below the ESC-50 state of the art of 99.1\% achieved by OmniVec2~\citep{srivastava2024omnivec2}, and also below the 94.7\% of PANNs CNN14 with fine-tuning~\citep{kong2020panns}. This gap is expected and informative rather than problematic: those systems leverage large-scale pretraining on AudioSet (approximately 2 million clips), whereas our model trains from scratch on just 1,600 samples per fold. The controlled comparison principle demands that both the \gls{snn} and \gls{ann} share identical data budgets, so the relevant benchmark is the \gls{ann} baseline, not the pretrained state of the art.

The inter-fold variance (standard deviation 3.07 percentage points) reflects genuine difficulty variation across ESC-50's predefined folds: Fold~3 is easiest (68.75\%), while Folds~4 and~5 are hardest (60.00\% and 60.25\% respectively). This fold-level variation persists across all models and encodings, confirming that it is a property of the data splits rather than training instability.

Human performance on ESC-50 is 81.3\%~\citep{piczak2015esc50}. Our baseline thus reaches 78.5\% of human accuracy, which is reasonable for a small model ($\sim$622K parameters) without pretraining, data augmentation, or ensemble methods.

% ==================================================================
\section{Spike Encoding Comparison}
\label{sec:encoding-results}
% ==================================================================

The central experiment of this thesis evaluates seven spike encoding methods under identical architecture, training protocol, and evaluation conditions. Table~\ref{tab:encoding-comparison} presents the complete per-fold results; Figure~\ref{fig:encoding-bar} visualises the means.

\begin{table}[t]
\centering
\caption{Five-fold cross-validation accuracy (\%) for all seven spike encoding methods using the SpikingCNN ($T = 25$ timesteps). Encodings are ordered by descending mean accuracy.}
\label{tab:encoding-comparison}
\small
\begin{tabular}{@{}lccccccc@{}}
\toprule
\textbf{Encoding} & \textbf{Fold 1} & \textbf{Fold 2} & \textbf{Fold 3} & \textbf{Fold 4} & \textbf{Fold 5} & \textbf{Mean} & \textbf{Std} \\
\midrule
Direct      & 40.50 & 48.50 & 48.25 & 54.00 & 44.50 & 47.15 & 4.50 \\
Rate        & 24.50 & 27.25 & 23.00 & 21.50 & 23.75 & 24.00 & 1.90 \\
Phase       & 22.50 & 22.25 & 25.00 & 24.25 & 26.75 & 24.15 & 1.66 \\
Population  & 22.75 & 18.50 & 15.75 & 22.00 & 16.75 & 19.15 & 2.79 \\
Latency     & 14.00 & 15.75 & 17.75 & 15.50 & 18.50 & 16.30 & 1.62 \\
Delta       &  8.25 &  7.75 &  7.25 &  7.50 &  5.50 &  7.25 & 0.94 \\
Burst       &  5.00 &  5.25 &  9.25 &  6.00 &  7.00 &  6.50 & 1.54 \\
\midrule
ConvANN     & 66.25 & 64.00 & 68.75 & 60.00 & 60.25 & 63.85 & 3.07 \\
\bottomrule
\end{tabular}
\end{table}

The encoding hierarchy is unambiguous: direct (47.15\%) $\gg$ rate (24.00\%) $\approx$ phase (24.15\%) $>$ population (19.15\%) $>$ latency (16.30\%) $\gg$ delta (7.25\%) $\approx$ burst (6.50\%). A paired $t$-test between the best \gls{snn} (direct) and the ConvANN baseline yields $t(4) = 8.64$, $p = 0.001$, confirming statistical significance.

\subsection{Information Preservation Principle}
\label{ssec:info-preservation}

The hierarchy is explained by a single organising principle: \textbf{encoding accuracy is proportional to the amount of input information preserved across the full temporal window}. Three factors determine information preservation:

\paragraph{Magnitude fidelity.}
Direct encoding feeds the continuous normalised spectrogram $\hat{s}_i \in [0, 1]$ at every timestep, preserving full magnitude information. Rate coding quantises intensity to stochastic binary spikes, discarding amplitude detail and introducing sampling noise. Phase and latency coding collapse magnitude into spike timing (one spike per neuron), retaining only $\mathcal{O}(\log T)$ bits per neuron versus the full floating-point precision of direct encoding.

\paragraph{Temporal coverage.}
Direct and rate coding provide input at all $T = 25$ timesteps, giving the LIF neurons a sustained driving current that enables membrane potentials to evolve through complex, information-rich trajectories. Phase coding distributes single spikes across the full $[1, T]$ window, achieving broad temporal coverage despite minimal spike count. Latency coding covers only a partial window, since high-intensity neurons fire early and the network receives diminishing input at later timesteps. Burst coding concentrates all spikes within the first 5 of 25 timesteps, leaving 80\% of the temporal window entirely silent---a catastrophic mismatch with LIF dynamics that require sustained integration.

\paragraph{Static-spectrogram compatibility.}
Delta coding is designed for dynamic inputs where temporal changes carry information. Applied to a static mel spectrogram (identical image repeated at each timestep), it detects only the small injected Gaussian noise ($\sigma = 0.02$), producing extremely sparse, uninformative spike trains. This fundamental incompatibility---applying a change-detection encoding to a static input---explains its near-chance performance.

\subsection{Rate--Phase Equivalence}
\label{ssec:rate-phase}

The most striking finding is that phase coding (24.15\%) matches rate coding (24.00\%) despite using dramatically fewer spikes: each active neuron fires exactly once under phase coding, compared to an average of approximately 12.5 spikes under rate coding (at mean intensity 0.5 over 25 timesteps). This equivalence demonstrates that \textbf{temporal coverage matters more than spike count}. Phase coding distributes its single spikes across the full $[1, T]$ window, providing broad temporal coverage that enables LIF neurons to integrate meaningful patterns. Rate coding achieves the same coverage through redundant stochastic spikes. The energy implication is significant: phase coding achieves equal accuracy at a fraction of the spike count, making it the most energy-efficient spike-based encoding.

\subsection{Population Coding Underperformance}
\label{ssec:population-underperformance}

Population coding (19.15\%) underperforms rate coding (24.00\%) despite using the same rate-coded input, because it replaces cross-entropy loss with MSE count loss over 500 output neurons (10 per class). The MSE loss landscape is harder to optimise: training accuracy reaches only 18--24\% at convergence, compared to 50--65\% for direct encoding. The expanded output layer also creates a high-dimensional target space where the network must learn to activate specific neuron subsets while suppressing others, a more difficult learning problem than the 50-way softmax used by other encodings.

% ==================================================================
\section{Rhythm-SNN and Advanced Techniques}
\label{sec:rhythm-results}
% ==================================================================

Beyond the seven standard encodings, we evaluated a suite of advanced techniques to improve \gls{snn} accuracy. Table~\ref{tab:advanced-techniques} presents the top ten configurations.

\begin{table}[t]
\centering
\caption{Top ten advanced SNN configurations on ESC-50 (five-fold cross-validation). The baseline SpikingCNN with direct encoding achieves 47.15\%. ``$\Delta$\,Baseline'' is the improvement over the direct-encoding baseline.}
\label{tab:advanced-techniques}
\small
\begin{tabular}{@{}lccl@{}}
\toprule
\textbf{Configuration} & \textbf{Mean (\%)} & \textbf{Std (\%)} & \textbf{$\Delta$\,Baseline (pp)} \\
\midrule
Dendritic + delays         & 61.65 & 3.69 & $+$14.50 \\
SR + Rhythm               & 61.35 & 3.54 & $+$14.20 \\
\textbf{Rhythm-SNN}       & \textbf{61.10} & \textbf{1.99} & $+$\textbf{13.95} \\
Rhythm + TET              & 60.75 & 2.16 & $+$13.60 \\
Rhythm + L1 ($10^{-5}$)   & 60.20 & 3.44 & $+$13.05 \\
Cochleagram SNN           & 55.35 & ---  & $+$8.20 \\
Stochastic resonance      & 50.40 & ---  & $+$3.25 \\
Learnable $\beta$         & 49.80 & ---  & $+$2.65 \\
ALIF (adaptive threshold) & 48.90 & ---  & $+$1.75 \\
TET loss                  & 48.50 & ---  & $+$1.35 \\
\bottomrule
\end{tabular}
\end{table}

The Rhythm-SNN achieves 61.10\% $\pm$ 1.99\%, a 13.95 percentage point improvement over the baseline \gls{snn} and only 2.75 percentage points below the \gls{ann} baseline of 63.85\%. Crucially, the Rhythm-SNN has the \emph{lowest} standard deviation (1.99\%) of all advanced configurations, indicating robust performance across folds. While dendritic computations with axonal delays achieve marginally higher mean accuracy (61.65\%), the Rhythm-SNN provides a better accuracy--stability trade-off and is selected as the primary model for SpiNNaker deployment.

The oscillatory modulation mechanism (Section~\ref{ssec:rhythm-snn}) works by creating temporal windows of heightened excitability. The learned frequencies and phases allow different neurons to become preferentially sensitive to different temporal features of the spectrogram, effectively implementing a form of temporal attention. This is biologically motivated by theta oscillations in mammalian auditory cortex~\citep{lakatos2005oscillatory}, which modulate neuronal excitability to align processing with rhythmic structure in auditory stimuli.

\paragraph{Negative results: knowledge distillation and data augmentation.}
Knowledge distillation from the ConvANN to the SpikingCNN consistently degraded accuracy across all configurations tested. The fundamental problem is that soft ANN logits encourage the SNN to produce smooth, graded outputs, which conflicts with the discrete spiking dynamics. Every combination with knowledge distillation performed worse than the equivalent configuration without it.

Data augmentation (time-shifting, frequency masking, Gaussian noise) also proved harmful: the augmented SNN achieved only 40.75\% $\pm$ 16.03\% (folds: 46.00, 48.75, 24.50, 63.75, 20.75\%)---6.40 percentage points \emph{worse} than the baseline, with variance tripled. Root cause analysis revealed that the early stopping criterion (patience = 10 epochs) terminates training prematurely on augmented data: Folds~3 and~5 stopped at epochs 39 and 33, well before convergence. The augmented \gls{ann} also degraded slightly (61.70\% $\pm$ 4.58\%, $-$2.15 pp). These negative results inform future work: augmentation requires careful interaction with early stopping when training data is limited.

% ==================================================================
\section{SpiNNaker Deployment}
\label{sec:spinnaker-results}
% ==================================================================

The Rhythm-SNN is deployed on SpiNNaker neuromorphic hardware using the current-injection hybrid pipeline described in Section~\ref{sec:spinnaker}. We evaluate two temporal configurations ($T = 3$ and $T = 1$) and a pruning sweep across 10 sparsity levels.

\subsection{Full FC1+FC2 Deployment}
\label{ssec:full-deploy-results}

Table~\ref{tab:spinnaker-t3} presents the per-fold results for the $T = 3$ full deployment, alongside the equivalent snnTorch simulation.

\begin{table}[t]
\centering
\caption{SpiNNaker full FC1+FC2 deployment accuracy (\%) at $T = 3$ timesteps. ``Gap'' is snnTorch accuracy minus SpiNNaker accuracy. ``Match'' is the percentage of samples classified identically by both platforms.}
\label{tab:spinnaker-t3}
\small
\begin{tabular}{@{}lcccccc@{}}
\toprule
\textbf{Fold} & \textbf{SpiNNaker} & \textbf{snnTorch} & \textbf{Gap (pp)} & \textbf{Match (\%)} & \textbf{Samples} & \textbf{Time (s)} \\
\midrule
1 & 55.00 & 56.50 & 1.50 & 84.25 & 400 & 1651 \\
2 & 60.50 & 63.00 & 2.50 & 82.75 & 400 & 1760 \\
3 & 55.50 & 58.00 & 2.50 & 83.00 & 400 & 1616 \\
4 & 60.00 & 61.00 & 1.00 & 86.00 & 400 & 1860 \\
5 & 55.75 & 59.00 & 3.25 & 80.75 & 400 & 1571 \\
\midrule
\textbf{Mean} & \textbf{57.35} & \textbf{59.50} & \textbf{2.15} & \textbf{83.35} & --- & --- \\
\textbf{Std}  & 2.39 & 2.38 & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

The SpiNNaker deployment achieves 57.35\% $\pm$ 2.39\% with a hardware--software gap of only 2.15 percentage points. This gap is remarkably small given the multiple sources of potential discrepancy: fixed-point versus floating-point arithmetic, exponential versus instantaneous synaptic currents, and discrete versus continuous time evolution. The 83.4\% match rate (proportion of samples classified identically by both platforms) demonstrates strong functional equivalence.

At $T = 1$ (single timestep), SpiNNaker achieves 55.05\% $\pm$ 1.96\% with a gap of 3.15 percentage points. The 2.30 percentage point drop from $T = 3$ to $T = 1$ reflects the reduced temporal integration available to the LIF neurons, but the model retains 96\% of its $T = 3$ accuracy, confirming that the Rhythm-SNN learns robust features that do not require extended temporal processing.

\subsection{Pruning Sweep on SpiNNaker}
\label{ssec:pruning-results}

Table~\ref{tab:pruning-sweep} presents the complete pruning sweep at $T = 3$ across all five folds and 10 sparsity levels, totalling 50 individual SpiNNaker deployments.

\begin{table}[t]
\centering
\caption{SpiNNaker pruning sweep ($T = 3$, five-fold mean). ``Gap'' = snnTorch $-$ SpiNNaker; negative gap means SpiNNaker outperforms snnTorch. Bold entries highlight cases where SpiNNaker matches or beats snnTorch.}
\label{tab:pruning-sweep}
\small
\begin{tabular}{@{}rccrc@{}}
\toprule
\textbf{Pruning (\%)} & \textbf{SpiNNaker (\%)} & \textbf{snnTorch (\%)} & \textbf{Gap (pp)} & \textbf{Std (\%)} \\
\midrule
0  (unpruned)  & 57.35 & 59.50 &  2.15 & 2.39 \\
50             & 60.45 & 61.85 &  1.40 & 1.98 \\
55             & 59.15 & 60.35 &  1.20 & 1.97 \\
\textbf{60}    & \textbf{60.15} & \textbf{59.70} & $\mathbf{-0.45}$ & 3.80 \\
65             & 59.60 & 60.40 &  0.80 & 3.12 \\
70             & 59.85 & 60.05 &  0.20 & 1.93 \\
75             & 59.65 & 60.60 &  0.95 & 3.11 \\
\textbf{80}    & \textbf{59.40} & \textbf{59.10} & $\mathbf{-0.30}$ & 2.62 \\
85             & 59.65 & 60.25 &  0.60 & 2.48 \\
90             & 58.05 & 58.40 &  0.35 & 3.03 \\
95             & 54.20 & 55.85 &  1.65 & 2.57 \\
\bottomrule
\end{tabular}
\end{table}

Three findings emerge from the pruning sweep. First, moderate pruning (50--85\%) \emph{improves} SpiNNaker accuracy relative to the unpruned model: 60.45\% at 50\% pruning versus 57.35\% unpruned. This counter-intuitive result is explained by the regularisation effect of pruning, which removes small-magnitude weights that contribute noise in fixed-point arithmetic. Second, at 60\% and 80\% pruning, SpiNNaker actually \emph{outperforms} snnTorch (gaps of $-$0.45 and $-$0.30 percentage points respectively), demonstrating that the hardware--software gap can be eliminated or even reversed through appropriate sparsification. Third, accuracy degrades gracefully: even at 95\% pruning (retaining only 5\% of weights), SpiNNaker achieves 54.20\%, only 3.15 percentage points below the unpruned deployment.

\paragraph{Poisson FC1 failure.}
An alternative deployment strategy using Poisson spike sources for FC1 input was attempted, where continuous feature values are converted to Poisson firing rates. This approach failed because the stochastic nature of Poisson spike generation introduces excessive variance in the FC1 input, particularly for the low-firing-rate neurons that carry discriminative information. The deterministic current-injection method provides reproducible inputs and was adopted as the standard deployment pipeline.

% ==================================================================
\section{Energy Analysis}
\label{sec:energy-results}
% ==================================================================

Energy per inference is estimated using the NeuroBench framework~\citep{yik2025neurobench} with the Horowitz cost model~\citep{horowitz2014energy}: 0.9\,pJ per \gls{ac} and 4.6\,pJ per \gls{mac}. Table~\ref{tab:energy} presents the energy analysis across temporal reduction and pruning configurations.

\begin{table}[t]
\centering
\caption{Estimated inference energy per sample under the NeuroBench framework (45\,nm CMOS). The ConvANN reference energy is 454\,nJ. ``Ratio'' is $E_{\text{ANN}} / E_{\text{SNN}}$, indicating the SNN's energy advantage over the ANN.}
\label{tab:energy}
\small
\begin{tabular}{@{}llcccc@{}}
\toprule
\textbf{Configuration} & \textbf{$T$} & \textbf{Energy (nJ)} & \textbf{Ratio vs ANN} & \textbf{Accuracy (\%)} \\
\midrule
SNN baseline ($T = 25$)     & 25 & 968  & $0.47\times$ & 47.15 \\
Rhythm-SNN ($T = 25$)       & 25 & 968  & $0.47\times$ & 61.10 \\
ConvANN                     & 1  & 454  & $1.0\times$  & 63.85 \\
\midrule
Rhythm-SNN ($T = 3$)        &  3 & 142  & $3.2\times$  & 59.50 \\
Rhythm-SNN ($T = 1$)        &  1 &  47  & $9.6\times$  & 58.20 \\
\midrule
85\% pruned ($T = 3$)       &  3 & $\sim$27 & $16.6\times$ & 59.65 \\
95\% pruned ($T = 3$)       &  3 & $\sim$14 & $32.4\times$ & 54.20 \\
\bottomrule
\end{tabular}
\end{table}

At $T = 25$, the \gls{snn} consumes 968\,nJ $\pm$ 37\,nJ per sample (five-fold validated), which is 2.1$\times$ \emph{more} expensive than the \gls{ann} (454\,nJ). The activation sparsity of 73.6\% $\pm$ 0.7\% corresponds to a spike rate of 26.4\%, well above the $\sim$6.4\% threshold identified by \citet{dampfhoffer2023snns} for software energy parity. This is a critical finding: \textbf{a standard SNN at full temporal resolution is not energy-efficient in software simulation}.

Temporal reduction transforms the energy landscape. At $T = 3$, energy drops to 142\,nJ---3.2$\times$ cheaper than the \gls{ann}---while maintaining 59.50\% accuracy (snnTorch). This is the first configuration where the SNN achieves a genuine energy advantage. At $T = 1$, energy falls further to 47\,nJ (9.6$\times$ advantage), with accuracy of 58.20\%.

Combining pruning with temporal reduction yields the most aggressive energy reduction. At 85\% pruning with $T = 3$, estimated energy is approximately 27\,nJ per sample---16.6$\times$ cheaper than the ConvANN---while SpiNNaker accuracy remains 59.65\%, essentially unchanged from the unpruned deployment. At 95\% pruning, energy drops to approximately 14\,nJ (32.4$\times$ advantage), though accuracy declines to 54.20\%.

The practical takeaway is that the Rhythm-SNN at $T = 3$ with 85\% pruning occupies the optimal point on the accuracy--energy Pareto frontier: it achieves 59.65\% accuracy at 27\,nJ, representing a 16.6$\times$ energy reduction relative to the \gls{ann} with a accuracy cost of only 4.20 percentage points.

% ==================================================================
\section{PANNs Transfer Learning}
\label{sec:panns-results}
% ==================================================================

The PANNs transfer experiment isolates the feature-learning bottleneck from spiking computation. Table~\ref{tab:panns} reports the five-fold results for all three classifier heads trained on frozen CNN14 embeddings.

\begin{table}[t]
\centering
\caption{PANNs transfer learning: classification accuracy (\%) of three classifier heads on frozen CNN14 embeddings (five-fold CV). The \gls{snn}--\gls{ann} gap collapses from 16.70\,pp (scratch) to 0.95\,pp (PANNs).}
\label{tab:panns}
\small
\begin{tabular}{@{}lccccccc@{}}
\toprule
\textbf{Head} & \textbf{Fold 1} & \textbf{Fold 2} & \textbf{Fold 3} & \textbf{Fold 4} & \textbf{Fold 5} & \textbf{Mean} & \textbf{Std} \\
\midrule
SNN     & 92.00 & 94.50 & 91.00 & 93.50 & 91.50 & 92.50 & 1.30 \\
ANN     & 93.00 & 95.00 & 92.00 & 95.50 & 91.75 & 93.45 & 1.54 \\
Linear  & 94.25 & 95.75 & 92.50 & 95.25 & 91.25 & 93.80 & 1.69 \\
\bottomrule
\end{tabular}
\end{table}

This is the most important finding of the thesis. The scratch-trained \gls{snn}--\gls{ann} gap of 16.70 percentage points (47.15\% versus 63.85\%) collapses to just 0.95 percentage points (92.50\% versus 93.45\%) when both architectures classify the same high-quality pretrained features. A paired $t$-test across folds yields $p = 0.034$, confirming that the residual gap is statistically significant but practically negligible.

This result demonstrates that the \gls{snn}--\gls{ann} gap is overwhelmingly a \textbf{feature-learning problem, not a spiking-computation problem}. When both models receive equally informative input representations, the spiking classifier performs comparably to the non-spiking one. The $\sim$622K-parameter scratch-trained \gls{snn} struggles not because LIF dynamics are inherently inferior classifiers, but because surrogate gradient training on 1,600 samples is insufficient to learn rich convolutional features---the same data limitation that also constrains the scratch-trained \gls{ann} to 63.85\% versus the pretrained 93.45\%.

The practical implication is immediate: a CNN14 feature extractor (running on a conventional processor) combined with a spiking classifier head (running on SpiNNaker) provides 92.50\% accuracy with neuromorphic inference energy. The feature extraction is performed once per sample at ingest time; the spiking classifier then operates at nanojoule-scale energy for all subsequent queries. This hybrid architecture is directly deployable.

A surprising secondary finding is that the linear head (93.80\%) marginally outperforms both the SNN head and the ANN head. This indicates that the CNN14 embeddings are already linearly separable for ESC-50's 50 classes, and the nonlinear hidden layers in the SNN and ANN heads provide no benefit---they can only introduce overfitting or optimisation noise on this particular downstream task.

% ==================================================================
\section{Adversarial Robustness}
\label{sec:adversarial-results}
% ==================================================================

The \gls{snn} and matched \gls{ann} are evaluated under \gls{fgsm}~\citep{goodfellow2015explaining} and \gls{pgd}~\citep{madry2018towards} attacks at multiple perturbation strengths. Table~\ref{tab:adversarial} reports the five-fold mean accuracy under attack, alongside the robustness ratio (SNN accuracy divided by ANN accuracy).

\begin{table}[t]
\centering
\caption{Adversarial robustness: five-fold mean accuracy (\%) under FGSM and PGD attacks at selected perturbation strengths $\varepsilon$. ``Ratio'' is SNN / ANN accuracy, indicating how many times more robust the SNN is.}
\label{tab:adversarial}
\small
\begin{tabular}{@{}llcccc@{}}
\toprule
\textbf{Attack} & $\boldsymbol{\varepsilon}$ & \textbf{SNN (\%)} & \textbf{ANN (\%)} & \textbf{SNN Std} & \textbf{Ratio} \\
\midrule
Clean ($\varepsilon = 0$) & 0 & 47.15 & 63.85 & 4.50 & $0.7\times$ \\
\midrule
FGSM & 0.01 & 37.50 & 24.15 & --- & $1.6\times$ \\
FGSM & 0.05 & 24.80 &  7.20 & --- & $3.4\times$ \\
FGSM & 0.10 & 16.55 &  2.75 & 5.49 & $6.0\times$ \\
FGSM & 0.20 &  8.10 &  2.05 & --- & $4.0\times$ \\
\midrule
PGD  & 0.01 & 30.45 & 11.60 & --- & $2.6\times$ \\
PGD  & 0.05 &  9.75 &  0.05 & --- & $195\times$ \\
\bottomrule
\end{tabular}
\end{table}

At \gls{fgsm} $\varepsilon = 0.1$, the \gls{snn} retains 16.55\% $\pm$ 5.49\% accuracy while the \gls{ann} collapses to 2.75\% $\pm$ 0.61\%, representing a 6.0$\times$ robustness advantage. Under the stronger \gls{pgd} attack at $\varepsilon = 0.05$, the disparity widens dramatically: the \gls{snn} retains 9.75\% while the \gls{ann} is reduced to 0.05\% (essentially random chance on 50 classes is 2\%), yielding a 195$\times$ robustness ratio.

The robustness advantage is explained by two mechanisms inherent to spiking computation. First, the binary thresholding of LIF neurons acts as a natural \emph{quantisation defence}: adversarial perturbations that modify continuous activations by small amounts are rounded away when the perturbed activation crosses or fails to cross a binary threshold. The perturbation must be large enough to flip the spike/no-spike decision, providing a built-in noise margin absent in continuous networks. Second, temporal integration over $T = 25$ timesteps provides an \emph{averaging effect}: a single-step adversarial perturbation influences the input at every timestep, but the LIF membrane dynamics integrate over the full trajectory, diluting the perturbation's impact relative to the accumulated clean signal~\citep{sharmin2020inherent}.

These results are validated across all five test folds (400 samples each, 2{,}000 total adversarial evaluations per $\varepsilon$ value), using the evaluation protocol recommended by \citet{wang2025reliable} for reliable \gls{snn} robustness assessment.

% ==================================================================
\section{Continual Learning}
\label{sec:cl-results}
% ==================================================================

Continual learning experiments evaluate catastrophic forgetting when the \gls{snn} and \gls{ann} are trained sequentially on ESC-50's five super-categories (Animals, Nature, Human, Domestic, Urban), each containing 10 classes. No replay buffer, regularisation, or architectural strategies are employed---both models face the ``bare'' sequential learning problem identically.
