% Chapter 2: Background and Related Work
% Included via \include{chapter2} from report.tex

\chapter{Background and Related Work}
\label{ch:background}

This chapter establishes the technical foundations on which this thesis is built. Section~\ref{sec:esc} introduces the environmental sound classification task and the ESC-50 benchmark. Section~\ref{sec:snn-background} formalises the spiking neural network framework, covering the \gls{lif} neuron model, surrogate gradient training, and the seven spike encoding methods evaluated in this work. Section~\ref{sec:neuromorphic-hardware} describes the SpiNNaker neuromorphic platform and situates it among contemporary alternatives. Section~\ref{sec:energy-background} examines the energy efficiency argument for \glspl{snn}, including the conditions under which spiking architectures genuinely outperform conventional networks. Section~\ref{sec:related-audio-snn} surveys prior work on \gls{snn}-based audio classification and identifies the gap this thesis addresses. Finally, Section~\ref{sec:adversarial-cl} reviews findings on adversarial robustness and continual learning in \glspl{snn}, motivating two of the advanced experiments in this report.

%=================================================================
\section{Environmental Sound Classification}
\label{sec:esc}
%=================================================================

Environmental sound classification (ESC) is a core task within computational auditory scene analysis, requiring a system to assign a categorical label to a short audio recording of a real-world acoustic event~\citep{barchiesi2015}. Applications include noise pollution monitoring in smart cities~\citep{bello2019}, automated wildlife census~\citep{stowell2019}, assistive technology for the hearing-impaired, and anomaly detection in industrial settings. Unlike speech recognition, ESC must handle the full diversity of non-linguistic acoustic events---from a dog barking to a chainsaw operating---without the benefit of structured phonological constraints.

\paragraph{The ESC-50 dataset.}
\gls{esc}~\citep{piczak2015} is the standard benchmark for this task. It comprises 2{,}000 five-second recordings drawn from the Freesound repository, distributed uniformly across 50 classes (40 clips per class). The classes span five broad super-categories: animals (e.g.\ dog, rooster, crow), natural soundscapes (e.g.\ rain, sea waves, thunderstorm), human non-speech sounds (e.g.\ clapping, coughing, laughing), domestic/indoor sounds (e.g.\ clock alarm, door knock, vacuum cleaner), and urban/outdoor sounds (e.g.\ helicopter, siren, car horn). Crucially, the dataset defines five predefined folds for cross-validation, ensuring that results are directly comparable across studies. Human classification performance on ESC-50 is 81.3\%~\citep{piczak2015}, reflecting the genuine perceptual difficulty of some distinctions (e.g.\ sea waves versus rain, insects versus crackling fire).

The current state of the art is 99.1\%, achieved by OmniVec2~\citep{srivastava2024} using a multi-modal Transformer pretrained on diverse data. Among audio-only methods, Pretrained Audio Neural Networks~(\gls{panns}) with CNN14 fine-tuning achieve 94.7\%~\citep{kong2020}, and Audio Spectrogram Transformers reach 95.6\%~\citep{gong2021}. These results rely on large-scale pretraining on AudioSet~\citep{gemmeke2017}, a dataset of approximately 2 million 10-second clips spanning 527 event categories. Without external pretraining, the best reported \gls{ann} performance on ESC-50 is substantially lower, motivating our choice to evaluate scratch-trained models as the primary comparison.

\paragraph{Input representation.}
The dominant input representation for ESC is the log-mel spectrogram, computed by applying the short-time Fourier transform to the audio waveform, mapping the resulting frequency bins through a bank of triangular mel-scale filters that approximate the frequency resolution of the human cochlea, and converting power to a logarithmic decibel scale. In this work, we extract spectrograms with 64 mel bands at a sampling rate of 22{,}050\,Hz and a hop length of 512 samples, yielding a representation of shape $(1, 64, 216)$ per clip---one channel, 64 frequency bins, and 216 time frames. Values are normalised to $[0, 1]$ per sample, a requirement for the spike encoding methods described in Section~\ref{subsec:encoding}.

\paragraph{Broader context.}
Other datasets complement ESC-50. UrbanSound8K~\citep{salamon2014} provides 8{,}732 clips across 10 urban sound classes; it is smaller in vocabulary and widely considered less challenging. AudioSet~\citep{gemmeke2017} provides massive scale (527 classes, 2 million clips) but uses weak labels and variable-quality recordings, making it unsuitable as a controlled evaluation benchmark. The Spiking Heidelberg Digits (SHD) and Spiking Speech Commands datasets~\citep{cramer2020} are natively spiking but address spoken-digit recognition rather than environmental sounds. No standard spiking benchmark exists for ESC, underscoring the need for this work.


%=================================================================
\section{Spiking Neural Networks}
\label{sec:snn-background}
%=================================================================

\Glspl{snn} are neural networks in which information is communicated through discrete, binary events called spikes. Unlike conventional \glspl{ann}, where neurons transmit continuous-valued activations, \gls{snn} neurons accumulate incoming signals over time in a membrane potential and emit a spike only when that potential exceeds a threshold. This event-driven, temporal computation paradigm is inspired by biological neural circuits and offers potential advantages in energy efficiency and temporal processing~\citep{maass1997,gerstner2002}.

%-------------------------------
\subsection{The Leaky Integrate-and-Fire Neuron}
\label{subsec:lif}
%-------------------------------

The \gls{lif} neuron is the most widely used model in modern \gls{snn} research, balancing biological plausibility with computational tractability~\citep{gerstner2002}. Its continuous-time dynamics describe an RC circuit with passive membrane leakage:

\begin{definition}[Leaky Integrate-and-Fire Neuron]
\label{def:lif}
The membrane potential $U(t)$ of a \gls{lif} neuron evolves according to the differential equation
\begin{equation}
    \tau_m \frac{dU(t)}{dt} = -U(t) + R \, I(t),
    \label{eq:lif-continuous}
\end{equation}
where $\tau_m$ is the membrane time constant, $R$ is the membrane resistance, and $I(t)$ is the input current. When $U(t)$ reaches the firing threshold $\vartheta$, the neuron emits a spike and the membrane potential resets:
\begin{equation}
    S(t) = \begin{cases} 1 & \text{if } U(t) \geq \vartheta, \\ 0 & \text{otherwise}, \end{cases}
    \qquad
    U(t^+) = U_{\mathrm{reset}} \quad \text{if } S(t) = 1.
    \label{eq:lif-spike}
\end{equation}
\end{definition}

For simulation in discrete timesteps of duration $\Delta t$, as required by GPU-based training frameworks, the continuous dynamics yield the recurrence:
\begin{equation}
    U[t+1] = \beta \cdot U[t] + W \cdot X[t] - S[t] \cdot \vartheta,
    \label{eq:lif-discrete}
\end{equation}
where $\beta = \exp(-\Delta t / \tau_m)$ is the decay factor, $W$ is the weight matrix, $X[t]$ is the input at timestep $t$, and the final term implements the post-spike reset by subtracting the threshold. The spike output is
\begin{equation}
    S[t] = \mathbf{1}\bigl[U[t] \geq \vartheta\bigr],
    \label{eq:spike-function}
\end{equation}
where $\mathbf{1}[\cdot]$ is the Heaviside step function. In this work, we set $\beta = 0.95$ (corresponding to $\tau_m \approx 20$\,ms at $\Delta t = 1$\,ms) and $\vartheta = 1.0$ across all layers, following the snnTorch framework defaults~\citep{eshraghian2023}.

The \gls{lif} model captures two essential properties of biological neurons that are absent from standard \gls{ann} units: \emph{temporal integration}, where evidence accumulates over multiple timesteps before a decision (spike) is made; and \emph{leaky memory}, where past inputs are exponentially forgotten, preventing indefinite charge build-up. These properties are central to the robustness advantages discussed in Section~\ref{sec:adversarial-cl}.

%-------------------------------
\subsection{Surrogate Gradient Training}
\label{subsec:surrogate}
%-------------------------------

The spike function in Equation~\eqref{eq:spike-function} is a Heaviside step, which has a derivative of zero almost everywhere and is undefined at the threshold. This renders standard backpropagation through time (BPTT) inapplicable, since the gradient $\partial S / \partial U$ vanishes for all sub-threshold and supra-threshold membrane potentials. For much of the history of \gls{snn} research, this non-differentiability confined learning to biologically inspired rules such as spike-timing-dependent plasticity (STDP)~\citep{bi1998}, which are local and unsupervised, and which do not scale to complex classification tasks.

The surrogate gradient method~\citep{neftci2019,zenke2021} resolves this by decoupling the forward and backward passes. The forward pass uses exact binary spikes as in Equation~\eqref{eq:spike-function}, preserving the discrete communication that underpins the energy argument. The backward pass substitutes a smooth, differentiable surrogate $\sigma(U - \vartheta)$ in place of the Heaviside:
\begin{equation}
    \frac{\partial \mathcal{L}}{\partial U} \approx \frac{\partial \mathcal{L}}{\partial S} \cdot \sigma'(U - \vartheta),
    \label{eq:surrogate-gradient}
\end{equation}
where $\sigma'$ is the derivative of the chosen surrogate function. This enables full BPTT through the unrolled spiking network, using the same optimisers, loss functions, and learning rate schedules as conventional deep learning.

Several surrogate functions have been proposed. The \emph{fast sigmoid} surrogate, used as the default in snnTorch~\citep{eshraghian2023}, is defined as $\sigma'(x) = 1 / (1 + k|x|)^2$ with slope parameter $k = 25$. The \emph{spike rate escape} surrogate uses an exponential form $\sigma'(x) = (1/\beta_s) \exp(-|x|/\beta_s)$, motivated by the theory of escape noise in stochastic neuron models~\citep{gerstner2002,gygax2025}. Other common choices include the arctangent, triangular, and straight-through estimator (STE) surrogates.

\citet{zenke2021} argued that the precise shape of the surrogate matters less than its slope, since all smooth approximations to the Heaviside convey similar gradient information. However, our ablation study in Chapter~\ref{ch:results} challenges this claim: across eight surrogates evaluated under identical conditions, we observe a bimodal distribution in which three surrogates (spike rate escape, fast sigmoid, arctangent) achieve 35--46\% accuracy while the remaining five (STE, sigmoid, sfs, triangular, and the stochastic variant) either fail to converge or collapse to chance performance. This suggests that surrogate selection is a consequential design decision, particularly for small-dataset regimes where training stability is fragile.

%-------------------------------
\subsection{Spike Encoding Methods}
\label{subsec:encoding}
%-------------------------------

Spike encoding is the process of converting continuous-valued inputs---in our case, mel spectrogram pixel intensities in $[0, 1]$---into temporal spike patterns suitable for \gls{snn} processing. The choice of encoding determines the information content per spike, the temporal structure of input activity, and the resulting energy cost. Seven methods are evaluated in this work, spanning the major categories identified in the computational neuroscience literature.

\paragraph{Rate coding~\citep{adrian1926}.}
Each input neuron emits a spike at each of the $T$ timesteps with probability equal to the input intensity:
\begin{equation}
    \mathrm{spk}[t, i] \sim \mathrm{Bernoulli}(x_i), \quad t = 1, \ldots, T.
    \label{eq:rate-coding}
\end{equation}
Rate coding is the oldest and most studied encoding in both neuroscience and \gls{snn} research. It provides maximal information content---$\mathcal{O}(T)$ bits per neuron---at the expense of high spike counts and corresponding energy cost. The stochastic nature introduces trial-to-trial variability, which can act as implicit regularisation but also limits reproducibility.

\paragraph{Latency coding~\citep{thorpe1996}.}
Each neuron fires at most once, with the spike time inversely related to input intensity:
\begin{equation}
    t_{\mathrm{fire}}(x_i) = \bigl\lceil (1 - x_i) \cdot (T - 1) \bigr\rceil.
    \label{eq:latency-coding}
\end{equation}
High-intensity inputs spike early, low-intensity inputs spike late or not at all. Information content is $\mathcal{O}(\log T)$ bits per neuron, and the single-spike constraint makes latency coding maximally energy-efficient. However, downstream \gls{lif} neurons must integrate over many timesteps before the late-arriving spikes contribute, creating a temporal bottleneck.

\paragraph{Delta coding~\citep{lichtsteiner2008}.}
Spikes encode temporal changes rather than absolute values:
\begin{equation}
    \mathrm{spk}[t, i] = \mathbf{1}\bigl[|x_i[t] - x_i[t-1]| > \delta\bigr],
    \label{eq:delta-coding}
\end{equation}
where $\delta$ is a threshold (set to 0.1 in this work). Delta coding is the computational analogue of dynamic vision sensors and retinal on/off cells. It is well suited to signals with rich temporal dynamics but performs poorly on static inputs: since a mel spectrogram is identical across all $T$ timesteps in the absence of temporal perturbation, delta coding produces near-zero spike activity, depriving downstream layers of information.

\paragraph{Direct (continuous) coding~\citep{rathi2021}.}
The normalised input is injected as a continuous current at every timestep:
\begin{equation}
    \mathrm{input}[t, i] = x_i, \quad \forall \, t = 1, \ldots, T.
    \label{eq:direct-coding}
\end{equation}
No explicit spike conversion is performed; the \gls{lif} neurons in the first hidden layer generate their own temporal spike patterns through the integrate-and-fire dynamics. This allows the network to learn an implicit, task-optimal encoding. Direct coding consistently achieves the highest accuracy in practice~\citep{rathi2021,eshraghian2023}, and is the best-performing encoding in our experiments (47.15\%).

\paragraph{Burst coding~\citep{izhikevich2004}.}
Each neuron emits a burst of spikes whose count is proportional to input intensity, concentrated at the start of the simulation window:
\begin{equation}
    n_{\mathrm{spikes}}(x_i) = \mathrm{round}(x_i \cdot N_{\max}), \qquad
    \mathrm{spk}[t, i] = \mathbf{1}\bigl[t < n_{\mathrm{spikes}}(x_i)\bigr],
    \label{eq:burst-coding}
\end{equation}
where $N_{\max} = 5$. This front-loads all spikes into the first $N_{\max}$ timesteps, leaving the remaining $T - N_{\max}$ timesteps silent. Biologically, burst firing is observed in auditory and visual cortex as a mechanism for reliable signal transmission through noisy synapses~\citep{lisman1997}.

\paragraph{Phase coding~\citep{okeefe1993}.}
Each neuron fires exactly once, at a time determined by mapping its intensity to a position within a single oscillation cycle:
\begin{equation}
    t_{\mathrm{fire}}(x_i) = \bigl\lfloor (1 - x_i) \cdot (T - 1) \bigr\rfloor, \quad x_i > 0.
    \label{eq:phase-coding}
\end{equation}
Phase coding is deterministic and produces exactly one spike per active neuron, making it the most energy-efficient single-spike scheme. It is inspired by theta-phase precession in hippocampal place cells~\citep{okeefe1993,montemurro2008}, where the phase of a spike relative to a background oscillation encodes spatial information.

\paragraph{Population coding~\citep{georgopoulos1986}.}
Rather than encoding information at the input, population coding modifies the output layer. Each of the 50 classes is represented by a population of $M = 10$ neurons, yielding 500 output units. Input uses rate coding, and the loss function is a mean squared error on spike counts:
\begin{equation}
    \mathcal{L} = \mathrm{MSE}\!\left(\frac{1}{T}\sum_{t=1}^{T} \mathbf{s}_{\mathrm{out}}[t], \;\mathbf{y}_{\mathrm{pop}}\right),
    \label{eq:population-coding}
\end{equation}
where $\mathbf{y}_{\mathrm{pop}} \in \{0,1\}^{500}$ is a binary target vector. Classification is by argmax of the total spike count summed within each class population.

To our knowledge, no prior work has systematically compared all seven encoding methods on a single audio benchmark under controlled conditions.


%=================================================================
\section{Neuromorphic Hardware}
\label{sec:neuromorphic-hardware}
%=================================================================

Neuromorphic hardware platforms implement spiking neuron dynamics directly in silicon, exploiting event-driven, parallel computation to achieve energy efficiency that is fundamentally unattainable with conventional von Neumann architectures~\citep{schuman2022}. This section describes SpiNNaker, the platform used in this thesis, and briefly surveys alternatives for context.

%-------------------------------
\subsection{SpiNNaker Architecture}
\label{subsec:spinnaker}
%-------------------------------

SpiNNaker (Spiking Neural Network Architecture) is a massively parallel neuromorphic platform developed at the University of Manchester~\citep{furber2014}. Each SpiNNaker chip contains 18 ARM968 processor cores, each capable of simulating up to 256 \gls{lif} neurons in real time at 1\,ms resolution. Chips are interconnected by a bespoke packet-switched network that routes binary spike events asynchronously via multicast: when a neuron spikes, a 4-byte packet containing its source address is broadcast to all connected post-synaptic populations. The network topology is a triangular torus, providing six bidirectional links per chip for scalable, fault-tolerant routing.

The full SpiNNaker machine comprises 1{,}036{,}800 ARM cores across 57{,}600 chips, capable of simulating up to one billion neurons in real time~\citep{furber2014}. For the experiments in this thesis, we use a SpiNN-5 board (48 chips, 864 cores) accessed remotely at \texttt{spinnaker.cs.man.ac.uk}.

Communication is exclusively through binary spikes---there is no mechanism for transmitting continuous values between populations. This imposes a hard architectural constraint: any network deployed on SpiNNaker must produce binary spike patterns at every inter-layer boundary. Weight values are stored in on-chip SDRAM and may be specified in fixed-point representation.

The sPyNNaker software stack~\citep{rhodes2018} provides a PyNN-compatible Python interface for defining populations, projections, and spike sources. Supported neuron models include \texttt{IF\_curr\_exp} (exponential current-based integrate-and-fire), \texttt{IF\_cond\_exp} (conductance-based), and the Izhikevich model. In this work, we use \texttt{IF\_curr\_exp} with carefully calibrated parameters (detailed in Chapter~\ref{ch:methodology}) to approximate the snnTorch \gls{lif} dynamics.

\paragraph{Constraints for deployment.}
Deploying a trained \gls{snn} on SpiNNaker entails three principal challenges. First, SpiNNaker neurons communicate exclusively through binary spikes, whereas operations such as average pooling in convolutional layers produce fractional activations. This fundamentally limits which layers can be deployed natively. Second, synaptic weights must be representable in the fixed-point format used by the ARM cores, requiring quantisation-aware transfer from floating-point trained models. Third, the 1\,ms simulation timestep and the maximum of 256 neurons per core impose constraints on network sizing and temporal resolution.

%-------------------------------
\subsection{Other Neuromorphic Platforms}
\label{subsec:other-neuromorphic}
%-------------------------------

Intel's Loihi~2~\citep{orchard2021} is a digital neuromorphic chip with 128 neuromorphic cores, each implementing up to 8{,}192 neurons with programmable synaptic dynamics and on-chip learning rules. Loihi~2 supports graded spikes (multi-bit events) in addition to binary spikes, and its programmable microcode architecture allows custom neuron models. Unlike SpiNNaker, Loihi~2 is a custom ASIC fabricated at an advanced process node, offering higher energy efficiency per operation but at reduced flexibility for large-scale reconfiguration.

BrainScaleS-2~\citep{pehle2022} takes a mixed-signal approach: analogue circuits implement neuron and synapse dynamics at physical timescales (approximately 1{,}000$\times$ faster than biological real time), while digital circuits handle configuration, calibration, and routing. The analogue substrate provides exceptional energy efficiency but introduces device mismatch that must be compensated through calibration.

IBM's TrueNorth~\citep{merolla2014} uses a digital, neurosynaptic architecture with one million neurons and 256 million synapses, consuming approximately 70\,mW at full utilisation. Its highly constrained neuron model (binary weights, no on-chip learning) limits deployment flexibility but enables extreme power efficiency.

It is important to note that SpiNNaker~1, used in this thesis, is a 130\,nm research platform designed for flexibility and programmability rather than energy minimality. Energy measurements from SpiNNaker are therefore conservative estimates; deployment on a modern neuromorphic ASIC such as Loihi~2 or SpiNNaker~2~\citep{mayr2019} would yield substantially lower figures.


%=================================================================
\section{Energy Efficiency in Spiking Neural Networks}
\label{sec:energy-background}
%=================================================================

The energy argument for \glspl{snn} rests on a distinction between two fundamental arithmetic operations. In a conventional \gls{ann}, each active connection requires a \gls{mac}: multiply the activation by the synaptic weight and accumulate the result. In an \gls{snn}, because spikes are binary (0 or 1), the multiplication reduces to a conditional \gls{ac}: if the pre-synaptic neuron spiked, add the weight; otherwise, do nothing.

\citet{horowitz2014} measured the energy cost of arithmetic operations at 45\,nm CMOS: a 32-bit floating-point \gls{mac} costs approximately 4.6\,pJ, while a 32-bit addition (\gls{ac}) costs approximately 0.9\,pJ. This 5.1$\times$ ratio is the foundation of the \gls{snn} energy advantage. However, this advantage is modulated by the spike rate: an \gls{snn} with $T$ timesteps and average spike rate $r$ performs $T \times r \times N_{\mathrm{syn}}$ \glspl{ac}, while a dense \gls{ann} performs $N_{\mathrm{syn}}$ \glspl{mac}. The \gls{snn} is cheaper when
\begin{equation}
    T \cdot r \cdot E_{\mathrm{AC}} < E_{\mathrm{MAC}},
    \label{eq:energy-crossover}
\end{equation}
which, substituting the Horowitz values, requires $T \cdot r < 5.1$.

\citet{dampfhoffer2023} formalised this analysis for software simulation, showing that \glspl{snn} must achieve a spike rate below approximately 6.4\% (for $T = 25$ timesteps at 4-bit quantisation) to outperform quantised \glspl{ann} in energy on conventional hardware. This is a demanding requirement: many trained \glspl{snn} exhibit spike rates of 20--40\%, well above the break-even point.

The picture changes on dedicated neuromorphic hardware. On platforms such as SpiNNaker or Loihi, each spike event is processed by dedicated routing and accumulation circuitry that consumes nanojoules per event, compared to the microjoule-scale cost of a GPU kernel launch. In this regime, the relevant comparison is not \gls{ac} versus \gls{mac} at the arithmetic level but rather the total platform energy per inference, which includes memory access, routing, and idle power. Neuromorphic platforms are designed to minimise all of these for sparse, event-driven workloads.

\paragraph{NeuroBench.}
The NeuroBench framework~\citep{yik2025}, published in \emph{Nature Communications}, provides a standardised methodology for benchmarking \gls{snn} efficiency. It defines three metrics: \emph{Effective ACs} (accumulate operations from binary activations), \emph{Effective MACs} (multiply-accumulate operations from non-binary activations), and \emph{Dense operations} (total operations assuming no sparsity). Energy is estimated as $E = N_{\mathrm{AC}} \times 0.9\,\text{pJ} + N_{\mathrm{MAC}} \times 4.6\,\text{pJ}$ at 45\,nm. This thesis adopts NeuroBench v2.2.0 for all energy comparisons, enabling direct comparison with other published results.


%=================================================================
\section{Related Work in SNN Audio Classification}
\label{sec:related-audio-snn}
%=================================================================

Despite the natural compatibility between temporal audio signals and spike-based computation, \gls{snn} research on audio classification remains sparse relative to the image domain.

\paragraph{SNN audio with environmental sounds.}
\citet{larroza2025} present the most directly relevant prior work: fully connected \glspl{snn} trained on ESC-10 (a 10-class subset of ESC-50) using rate, latency, and direct encoding. They report approximately 60\% accuracy with direct encoding. However, their study has three key limitations that this thesis addresses: it uses only fully connected architectures (no convolution), it evaluates only 10 classes (not the full 50-class benchmark), and it includes no hardware deployment. Our work extends to convolutional \glspl{snn} on the full ESC-50, with four additional encoding methods and deployment on SpiNNaker.

\citet{dominguezmorales2016} deployed a multilayer \gls{snn} on SpiNNaker for audio classification, but used synthetic pure tones (130--1{,}397\,Hz) rather than real-world recordings. While their work demonstrated the feasibility of \gls{snn}-based audio processing on neuromorphic hardware, the task difficulty is not comparable to ESC-50, which requires distinguishing 50 categories of complex, naturally occurring sounds.

\paragraph{SNN speech recognition.}
In the speech domain, the Spiking Heidelberg Digits (SHD) dataset~\citep{cramer2020} has become a standard benchmark for recurrent \glspl{snn}, with recent architectures achieving above 90\% accuracy. Keyword spotting on Google Speech Commands has been addressed with \glspl{snn} achieving competitive accuracy at reduced energy~\citep{pellegrini2021}. However, speech tasks benefit from the structured, quasi-periodic nature of vocal production, which provides temporal regularities absent in environmental sounds.

\paragraph{SNN image classification for context.}
The most mature application of deep \glspl{snn} is image classification. On CIFAR-10, state-of-the-art \glspl{snn} trained with surrogate gradients achieve 90--93\% accuracy~\citep{fang2021,zheng2021}, compared to 95\%+ for equivalent \glspl{ann}. This 2--5 percentage point gap is characteristic of the accuracy cost of spiking computation at the current state of the art. On ImageNet, \glspl{snn} have reached approximately 75\% top-1 accuracy with deep residual architectures~\citep{hu2024}. These results provide a useful reference for interpreting the \gls{snn}-\gls{ann} accuracy gap observed in our experiments.

\paragraph{The gap this thesis fills.}
Table~\ref{tab:related-work} summarises the landscape. No published work evaluates convolutional \glspl{snn} on the full 50-class ESC-50 benchmark, no work provides a systematic comparison of more than three spike encoding methods for audio, and no work deploys an \gls{snn} for environmental sound classification on neuromorphic hardware. This thesis addresses all three gaps simultaneously.

\begin{table}[t]
    \centering
    \caption{Summary of related SNN audio classification work. ``Conv'' indicates whether convolutional layers are used. ``HW'' indicates neuromorphic hardware deployment.}
    \label{tab:related-work}
    \small
    \begin{tabular}{@{}lllccc@{}}
        \toprule
        \textbf{Study} & \textbf{Task} & \textbf{Encodings} & \textbf{Conv} & \textbf{HW} & \textbf{Acc.} \\
        \midrule
        \citet{dominguezmorales2016} & Pure tones & Rate & No & SpiNNaker & $\sim$90\% \\
        \citet{cramer2020}           & SHD digits & Native spikes & No & --- & 83\% \\
        \citet{pellegrini2021}       & Keywords   & Direct & Yes & --- & 94\% \\
        \citet{larroza2025}          & ESC-10     & Rate, lat., dir. & No & --- & $\sim$60\% \\
        \midrule
        \textbf{This work}           & \textbf{ESC-50} & \textbf{7 methods} & \textbf{Yes} & \textbf{SpiNNaker} & \textbf{47.15\%} \\
        \bottomrule
    \end{tabular}
\end{table}


%=================================================================
\section{Adversarial Robustness and Continual Learning}
\label{sec:adversarial-cl}
%=================================================================

Two additional properties of \glspl{snn}---adversarial robustness and resistance to catastrophic forgetting---have been hypothesised in the literature and are evaluated empirically in this thesis.

%-------------------------------
\subsection{Adversarial Robustness}
\label{subsec:adversarial-background}
%-------------------------------

Adversarial examples are inputs crafted to maximise classification error through small, often imperceptible perturbations~\citep{goodfellow2015}. The Fast Gradient Sign Method (\gls{fgsm}) generates adversarial examples as
\begin{equation}
    \mathbf{x}' = \mathbf{x} + \varepsilon \cdot \mathrm{sign}\!\left(\nabla_{\mathbf{x}} \mathcal{L}(\mathbf{x}, y)\right),
    \label{eq:fgsm}
\end{equation}
where $\varepsilon$ controls the perturbation magnitude. \Gls{pgd}~\citep{madry2018} extends this to an iterative, multi-step attack that is generally stronger.

\citet{sharmin2020} demonstrated that \glspl{snn} with rate encoding exhibit higher adversarial accuracy than equivalent \glspl{ann} in black-box transfer attack scenarios. They attributed this robustness to two mechanisms. First, the firing threshold acts as a nonlinear filter: small perturbations to the membrane potential may not be sufficient to cross the threshold, leaving the spike pattern unchanged. Second, the stochasticity of rate coding means that the same perturbation produces different spike patterns across trials, disrupting gradient-based attacks that assume deterministic input-output mappings.

A critical methodological caveat is raised by \citet{wang2025sapgd}, who show that standard \gls{pgd} \emph{underestimates} the true adversarial robustness of \glspl{snn}. The vanishing surrogate gradients in deep spiking networks weaken the attack itself, making the \gls{snn} appear more robust than it is. They propose Spike-Aware PGD (SA-PGD), which compensates for gradient attenuation. This finding is important context for interpreting our adversarial results in Chapter~\ref{ch:results}: we use standard \gls{pgd} following \citet{sharmin2020} but acknowledge that our robustness figures may be conservative upper bounds on the true \gls{snn} advantage.

An additional mechanism specific to our architecture is \emph{gradient masking} through the spike nonlinearity. The derivative of the spike function is zero almost everywhere (Equation~\ref{eq:spike-function}), and the surrogate gradient used during training is only an approximation. Adversarial attacks that rely on accurate gradient computation are therefore inherently weakened when applied to \glspl{snn}, regardless of any genuine robustness advantage. Disentangling true robustness from gradient masking remains an open problem~\citep{wang2025sapgd}.

%-------------------------------
\subsection{Continual Learning and Catastrophic Forgetting}
\label{subsec:cl-background}
%-------------------------------

Catastrophic forgetting~\citep{mccloskey1989,french1999} occurs when a neural network, trained sequentially on multiple tasks, loses performance on earlier tasks as it adapts to later ones. This is a fundamental limitation of gradient-based learning with shared parameters, since weight updates that optimise for the current task can overwrite representations critical for previous tasks.

Proposed mitigations include elastic weight consolidation~\citep{kirkpatrick2017}, progressive neural networks~\citep{rusu2016}, and experience replay~\citep{rolnick2019}, each adding complexity and memory overhead. An orthogonal hypothesis is that \glspl{snn} may be inherently less susceptible to forgetting due to their sparse, temporally distributed activation patterns. When only a small fraction of neurons are active for any given input (high activation sparsity), weight updates are naturally confined to a sparse subset of parameters, reducing interference with representations for other tasks.

\citet{golden2022} provided evidence for this hypothesis by showing that offline consolidation---an analogue of sleep-based memory replay---prevents catastrophic forgetting in \glspl{snn} without explicit regularisation. \citet{zhang2023} introduced NACA (Neuronal Activation-based Credit Assignment) for online \gls{snn} continual learning, achieving competitive performance on split-MNIST and split-CIFAR benchmarks.

In this thesis, we evaluate catastrophic forgetting under the simplest and most stringent protocol: sequential fine-tuning on five ESC-50 super-categories (10 classes each) without replay, regularisation, or architectural expansion. This isolates the inherent forgetting properties of the \gls{snn} and \gls{ann} architectures, removing confounds introduced by anti-forgetting mechanisms. Backward Transfer (BWT) is used as the primary metric:
\begin{equation}
    \mathrm{BWT} = \frac{1}{K-1}\sum_{i=1}^{K-1}\bigl(R_{K,i} - R_{i,i}\bigr),
    \label{eq:bwt}
\end{equation}
where $R_{K,i}$ is the accuracy on task $i$ after training on all $K$ tasks, and $R_{i,i}$ is the accuracy on task $i$ immediately after training on it. Negative BWT indicates forgetting.


%=================================================================
\section{Summary}
\label{sec:bg-summary}
%=================================================================

This chapter has established the technical context for the thesis. The ESC-50 dataset provides a rigorous, standardised benchmark for environmental sound classification with well-understood human and machine baselines. The \gls{lif} neuron model and surrogate gradient training form the computational foundation for modern deep \glspl{snn}. Seven spike encoding methods offer distinct trade-offs between information content, energy cost, and biological plausibility. SpiNNaker provides a concrete neuromorphic target for deployment, with binary spike communication as both its defining advantage and its principal constraint. The energy efficiency of \glspl{snn} is contingent on spike rate and execution platform, with NeuroBench providing a standardised measurement framework. Finally, adversarial robustness and continual learning represent two under-explored advantages of spiking computation that are directly testable.

The critical observation from this literature review is that \emph{no prior work has evaluated convolutional \glspl{snn} on the full ESC-50 benchmark, systematically compared more than three encoding methods for audio, or deployed an \gls{snn} for environmental sound classification on neuromorphic hardware}. The following chapter describes the methodology designed to address these gaps.
