\chapter{Methodology}
\label{ch:methodology}

This chapter describes the experimental design, dataset preprocessing, network architecture, spike encoding implementations, training protocol, SpiNNaker deployment pipeline, energy measurement framework, and advanced evaluation methods used throughout this work. Every design decision is motivated by the principle of controlled comparison: the SNN and ANN differ \emph{only} in their neuron model, so that observed performance differences are attributable to the spiking formalism rather than to confounding architectural or training factors.

% ==================================================================
\section{Design Philosophy}
\label{sec:design-philosophy}

The central experimental principle is \textbf{controlled comparison}. Both the \gls{snn} and \gls{ann} share identical convolutional feature extraction layers, identical fully connected classifier dimensions, and identical training hyperparameters. They are trained with the same optimiser, learning rate schedule, early stopping criterion, and evaluated under the same five-fold cross-validation protocol. The sole difference is the neuron activation: the SNN uses \gls{lif} neurons with surrogate gradient training~\citep{neftci2019surrogate}, whereas the ANN uses ReLU activations with standard backpropagation.

This design follows the recommendation of \citet{deng2021optimal} for fair SNN--ANN comparison and ensures that any accuracy gap can be attributed to the spiking formalism and spike encoding scheme, not to differences in model capacity, regularisation, or data exposure. The parameter count is deliberately kept small ($\sim$622K) to avoid overfitting on the 1,600 training samples available per fold in ESC-50, and to remain deployable on SpiNNaker hardware within its per-core memory constraints.

% ==================================================================
\section{Dataset and Preprocessing}
\label{sec:dataset}

\paragraph{ESC-50.}
ESC-50~\citep{piczak2015esc} comprises 2,000 five-second environmental sound recordings drawn from Freesound.org, evenly distributed across 50 classes with 40 clips per class. The classes are grouped into five broad super-categories: \emph{Animals} (classes 0--9), \emph{Natural soundscapes} (10--19), \emph{Human non-speech sounds} (20--29), \emph{Domestic sounds} (30--39), and \emph{Urban noises} (40--49). The dataset provides five predefined cross-validation folds that are balanced by class and avoid speaker or recording overlap across folds; we use these throughout. Human classification accuracy on ESC-50 is 81.3\%~\citep{piczak2015esc}.

\paragraph{Preprocessing.}
All audio clips are loaded at a sample rate of 22,050\,Hz and zero-padded to exactly 110,250 samples (5.0\,s). A log-mel spectrogram is then computed:
\begin{equation}
\label{eq:mel}
\mathbf{S} = 10 \cdot \log_{10}\!\bigl(\text{MelFilterbank}(|\text{STFT}(y)|^2) \,/\, S_{\max}\bigr),
\end{equation}
where the STFT uses $n_\text{fft} = 1024$ and hop length $= 512$, and the mel filterbank has $n_\text{mels} = 64$ bins spanning the full frequency range (0--11,025\,Hz). The resulting spectrogram has shape $(64 \times 216)$.

Each spectrogram is then min--max normalised per sample:
\begin{equation}
\label{eq:norm}
\hat{s}_{ij} = \frac{s_{ij} - \min(\mathbf{S})}{\max(\mathbf{S}) - \min(\mathbf{S})},
\end{equation}
yielding values in $[0, 1]$. This normalisation is essential for spike encoding: rate coding generates spikes with probability equal to pixel intensity, latency coding maps intensity to spike time, and direct coding feeds values as membrane current---all require a bounded $[0, 1]$ input range. A channel dimension is prepended to produce the final input tensor of shape $(1, 64, 216)$.

All 2,000 spectrograms are precomputed and cached in memory prior to training. This eliminates I/O overhead and ensures deterministic data loading with \texttt{librosa}~\citep{mcfee2015librosa}.

% ==================================================================
\section{Network Architecture}
\label{sec:architecture}

Both the SNN and ANN follow the same four-layer convolutional--fully-connected topology, shown in Figure~\ref{fig:architecture} and summarised in Table~\ref{tab:architecture}. The shared structure ensures that any performance difference between the two models can be attributed solely to the neuron model.

\subsection{SpikingCNN}
\label{ssec:snn-architecture}

The SNN processes input as a sequence of $T = 25$ timesteps. At each timestep $t$, the encoded input $\mathbf{x}^{(t)} \in \mathbb{R}^{B \times 1 \times 64 \times 216}$ passes through two convolutional blocks and two fully connected layers:
\begin{enumerate}[nosep]
    \item \textbf{Block~1:} Conv2d$(1 \to 32$, $k\!=\!3$, $p\!=\!1)$ $\to$ BatchNorm2d(32) $\to$ MaxPool2d(2) $\to$ LIF$_1$
    \item \textbf{Block~2:} Conv2d$(32 \to 64$, $k\!=\!3$, $p\!=\!1)$ $\to$ BatchNorm2d(64) $\to$ MaxPool2d(2) $\to$ LIF$_2$
    \item \textbf{Pooling:} AvgPool2d$(4, 6)$ $\to$ flatten to $\mathbb{R}^{2304}$
    \item \textbf{FC~1:} Linear$(2304 \to 256)$ $\to$ LIF$_3$
    \item \textbf{FC~2:} Linear$(256 \to 50)$ $\to$ LIF$_4$
\end{enumerate}

Each \gls{lif} neuron implements the discrete-time leaky integrate-and-fire dynamics:
\begin{equation}
\label{eq:lif}
U_i^{(t)} = \beta \, U_i^{(t-1)} + I_i^{(t)}, \quad
S_i^{(t)} = \Theta\!\bigl(U_i^{(t)} - \vartheta\bigr), \quad
U_i^{(t)} \leftarrow U_i^{(t)}\bigl(1 - S_i^{(t)}\bigr),
\end{equation}
where $U_i^{(t)}$ is the membrane potential, $I_i^{(t)}$ is the synaptic input current, $\beta = 0.95$ is the decay factor (membrane time constant), $\vartheta = 1.0$ is the spike threshold, $\Theta(\cdot)$ is the Heaviside step function, and the final term implements soft reset (membrane is reset to zero upon spiking). The non-differentiable $\Theta$ is replaced during backpropagation by the fast sigmoid surrogate gradient~\citep{zenke2021remarkable}:
\begin{equation}
\label{eq:surrogate}
\frac{\partial S}{\partial U} \approx \frac{1}{(1 + \alpha\,|U - \vartheta|)^2}, \quad \alpha = 25.
\end{equation}

The AvgPool2d$(4, 6)$ layer reduces the feature map from $(16, 54)$ to $(4, 9)$ after two rounds of MaxPool2d(2). We use AvgPool2d with explicit kernel size rather than AdaptiveAvgPool2d because PyTorch's adaptive pooling lacks Apple MPS backend support, and our development workflow requires MPS compatibility for rapid iteration before deploying to NVIDIA A100 GPUs on the CSF3 cluster.

The model outputs both spikes $\mathbf{S}^{(t)} \in \{0, 1\}^{B \times 50}$ and membrane potentials $\mathbf{U}^{(t)} \in \mathbb{R}^{B \times 50}$ at each of the 25 timesteps. At inference, the predicted class is:
\begin{equation}
\label{eq:prediction}
\hat{y} = \arg\max_c \sum_{t=1}^{T} U_c^{(t)}.
\end{equation}

\subsection{ConvANN Baseline}
\label{ssec:ann-architecture}

The ANN baseline mirrors the SNN exactly, replacing each LIF neuron with a ReLU activation and removing the temporal dimension. A single addition is Dropout$(p = 0.3)$ before the final fully connected layer, which compensates for the ANN's lack of intrinsic stochastic regularisation (the SNN's binary spiking provides implicit noise regularisation). The ANN processes a single $(1, 64, 216)$ input and outputs logits $\in \mathbb{R}^{50}$ in a single forward pass.

\begin{table}[t]
\centering
\caption{Layer-by-layer comparison of the SpikingCNN and ConvANN architectures. Both models have approximately 622K trainable parameters. The SNN processes $T = 25$ timesteps; the ANN processes a single frame.}
\label{tab:architecture}
\small
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Layer} & \textbf{SNN (SpikingCNN)} & \textbf{ANN (ConvANN)} & \textbf{Output Shape} \\
\midrule
Input & Encoded: $(T, B, 1, 64, 216)$ & Raw: $(B, 1, 64, 216)$ & --- \\
Conv1 & Conv2d$(1\!\to\!32, 3)$ + BN & Conv2d$(1\!\to\!32, 3)$ + BN & $(32, 64, 216)$ \\
Act1 + Pool & MaxPool(2) $\to$ LIF$(\beta\!=\!0.95)$ & ReLU $\to$ MaxPool(2) & $(32, 32, 108)$ \\
Conv2 & Conv2d$(32\!\to\!64, 3)$ + BN & Conv2d$(32\!\to\!64, 3)$ + BN & $(64, 32, 108)$ \\
Act2 + Pool & MaxPool(2) $\to$ LIF$(\beta\!=\!0.95)$ & ReLU $\to$ MaxPool(2) & $(64, 16, 54)$ \\
AvgPool & AvgPool2d$(4, 6)$ & AvgPool2d$(4, 6)$ & $(64, 4, 9)$ \\
Flatten & --- & --- & $2304$ \\
FC1 & Linear$(2304\!\to\!256)$ + LIF & Linear$(2304\!\to\!256)$ + ReLU & $256$ \\
Dropout & --- & Dropout$(0.3)$ & $256$ \\
FC2 & Linear$(256\!\to\!50)$ + LIF & Linear$(256\!\to\!50)$ & $50$ \\
\midrule
Parameters & 621,938 & 621,938 & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Rhythm-SNN: Oscillatory Threshold Modulation}
\label{ssec:rhythm-snn}

Motivated by theta oscillations observed in mammalian auditory cortex~\citep{lakatos2005oscillatory}, we introduce a variant that modulates the effective firing threshold with a learnable sinusoidal signal. The Rhythm-SNN replaces each standard LIF neuron with a \emph{RhythmLIF} neuron:
\begin{equation}
\label{eq:rhythm}
U_i^{(t)} = \beta_i \, U_i^{(t-1)} + I_i^{(t)} + A_i \sin\!\left(\frac{2\pi f_i \, t}{T} + \phi_i\right),
\end{equation}
where $A_i$, $f_i$, and $\phi_i$ are per-neuron learnable amplitude, frequency, and phase parameters respectively. The decay factor $\beta_i$ and threshold $\vartheta_i$ are also learnable (parameterised through a sigmoid to ensure $\beta_i \in (0, 1)$). The oscillatory term acts as a subthreshold current that creates temporal windows of heightened excitability, enabling the network to develop rhythmic processing patterns.

At initialisation, $A_i = 0.1$, $f_i \sim \mathcal{U}(0.5, 5.0)$, and $\phi_i = 0$. These parameters are trained end-to-end with the synaptic weights via surrogate gradient descent. The Rhythm-SNN achieves 61.10\% $\pm$ 1.99\%, a 13.95 percentage point improvement over the standard SNN and only 2.75 percentage points below the ANN baseline.

% ==================================================================
\section{Spike Encoding Implementations}
\label{sec:encodings}

Seven encoding methods are evaluated, transforming the normalised spectrogram $\hat{\mathbf{S}} \in [0, 1]^{1 \times 64 \times 216}$ into a temporal input tensor $\mathbf{X} \in \mathbb{R}^{T \times 1 \times 64 \times 216}$ with $T = 25$ timesteps. All encodings are implemented using snnTorch~\citep{eshraghian2023training} and custom PyTorch code.

\paragraph{Rate coding.}
Each input pixel generates independent Bernoulli spikes at each timestep with probability equal to its intensity:
\begin{equation}
\label{eq:rate}
X_i^{(t)} \sim \text{Bernoulli}(\hat{s}_i), \quad t = 1, \ldots, T.
\end{equation}
Information content: $\mathcal{O}(T)$ bits per neuron. Stochastic.

\paragraph{Latency coding.}
Each neuron fires exactly once; higher intensity maps to an earlier spike time:
\begin{equation}
\label{eq:latency}
t_{\text{fire}}(\hat{s}_i) = \left\lfloor (1 - \hat{s}_i) \cdot (T - 1) \right\rfloor, \quad \hat{s}_i > \epsilon,
\end{equation}
using a linearised approximation with $\tau = 5.0$ and minimum threshold $\epsilon = 0.01$. Neurons with intensity below $\epsilon$ remain silent. Information content: $\mathcal{O}(\log T)$ bits per neuron. Deterministic.

\paragraph{Delta coding.}
Spikes are emitted on positive temporal changes exceeding a threshold $\theta_\delta = 0.1$. Since the spectrogram is inherently static, we inject small Gaussian noise ($\sigma = 0.02$) at each timestep to create temporal variation:
\begin{equation}
\label{eq:delta}
X_i^{(t)} = \Theta\!\bigl((\hat{s}_i + \mathcal{N}(0, \sigma^2))^{(t)} - (\hat{s}_i + \mathcal{N}(0, \sigma^2))^{(t-1)} - \theta_\delta\bigr).
\end{equation}
Biologically motivated by on-centre cells in the auditory pathway, but poorly suited to static inputs. Stochastic.

\paragraph{Direct (continuous) coding.}
The normalised spectrogram is repeated unchanged at each timestep, providing continuous current input to the LIF neurons:
\begin{equation}
\label{eq:direct}
X_i^{(t)} = \hat{s}_i, \quad \forall\, t = 1, \ldots, T.
\end{equation}
No explicit spike conversion occurs at the input; the LIF neurons generate their own spike timing through membrane integration. This allows the network to learn an implicit, task-optimal temporal code. Deterministic.

\paragraph{Burst coding.}
Spike count is proportional to intensity, concentrated in the earliest timesteps:
\begin{equation}
\label{eq:burst}
n_i = \text{round}(\hat{s}_i \times N_{\max}), \quad X_i^{(t)} = \mathbf{1}[t < n_i],
\end{equation}
with $N_{\max} = 5$, giving a maximum spike density of 5/25 = 20\%. Biologically motivated by burst-firing neurons in the auditory cortex. Deterministic.

\paragraph{Phase coding.}
Intensity is mapped to spike timing within a single oscillation period spanning $T$ timesteps:
\begin{equation}
\label{eq:phase}
t_{\text{fire}}(\hat{s}_i) = \left\lfloor (1 - \hat{s}_i) \cdot (T - 1) \right\rfloor, \quad \hat{s}_i > 0.
\end{equation}
High intensity fires early (small phase offset); each active neuron fires exactly once. Neurons with zero intensity are silent. Biologically motivated by theta-phase precession in the hippocampus. Deterministic.

\paragraph{Population coding.}
The output layer is expanded to 500 neurons (10 per class), with rate-coded input. The loss function is MSE count loss:
\begin{equation}
\label{eq:population}
\mathcal{L}_{\text{pop}} = \left\lVert \frac{1}{T}\sum_{t=1}^{T} \mathbf{S}_{\text{out}}^{(t)} - \mathbf{y}_{\text{pop}} \right\rVert^2,
\end{equation}
where $\mathbf{y}_{\text{pop}} \in \{0, 1\}^{500}$ is the target vector with 1s for the 10 neurons of the correct class. Classification uses the argmax of grouped spike counts.

Table~\ref{tab:encodings} summarises the properties of all seven encodings.

\begin{table}[t]
\centering
\caption{Summary of the seven spike encoding methods. ``Spikes/neuron'' indicates the expected number of spikes per active input neuron over $T = 25$ timesteps for a typical input intensity of 0.5.}
\label{tab:encodings}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Encoding} & \textbf{Type} & \textbf{Spikes/neuron} & \textbf{Temporal coverage} & \textbf{Accuracy (\%)} \\
\midrule
Direct     & Deterministic & Implicit (LIF) & Full $[1, T]$ & 47.15 \\
Rate       & Stochastic    & $\sim$12.5      & Full $[1, T]$ & 24.00 \\
Phase      & Deterministic & 1               & Full $[1, T]$ & 24.15 \\
Population & Stochastic    & $\sim$12.5 (input) & Full $[1, T]$ & 19.15 \\
Latency    & Deterministic & 1               & Partial        & 16.30 \\
Delta      & Stochastic    & $\sim$0.5       & Sparse         & 7.25 \\
Burst      & Deterministic & 2.5             & First $[1, 5]$ only & 6.50 \\
\bottomrule
\end{tabular}
\end{table}

% ==================================================================
\section{Training Protocol}
\label{sec:training}

Both models are trained using the Adam optimiser~\citep{kingma2015adam} with an initial learning rate of $10^{-3}$ and weight decay of $10^{-4}$. The learning rate is reduced on plateau by a factor of 0.5 with patience of 5 epochs, monitoring validation loss. Early stopping with patience of 10 epochs prevents overfitting on the small training set (1,600 samples per fold). The maximum number of epochs is 50 and the batch size is 32.

\paragraph{SNN loss.}
The SNN is trained with a per-timestep cross-entropy loss summed over all $T$ timesteps~\citep{eshraghian2023training}:
\begin{equation}
\label{eq:snn-loss}
\mathcal{L}_{\text{SNN}} = \sum_{t=1}^{T} \text{CE}\!\left(\mathbf{U}^{(t)},\, y\right),
\end{equation}
where $\mathbf{U}^{(t)} \in \mathbb{R}^{B \times 50}$ is the output membrane potential at timestep $t$ and $y$ is the ground truth label. This formulation provides gradient flow through all timesteps, yielding better convergence than rate-based loss functions that aggregate only the final output~\citep{eshraghian2023training}.

\paragraph{ANN loss.}
The ANN uses standard cross-entropy on the single-pass logits: $\mathcal{L}_{\text{ANN}} = \text{CE}(\mathbf{z}, y)$.

\paragraph{Evaluation.}
Five-fold cross-validation follows the ESC-50 predefined splits: in each fold, four folds form the training set and one fold is held out for testing. The reported metric is the best validation accuracy achieved at any epoch (not the final-epoch accuracy). All reported means and standard deviations are computed across the five test folds.

\paragraph{Computational resources.}
All five-fold cross-validation experiments are run on the University of Manchester CSF3 high-performance computing cluster using NVIDIA A100 SXM4 80\,GB GPUs. Training one fold of the SNN takes approximately 6 minutes on an A100, compared to 65 minutes on Apple M-series MPS---an order-of-magnitude speedup. CUDA 12.6 and PyTorch 2.10 are used throughout.

% ==================================================================
\section{SpiNNaker Deployment}
\label{sec:spinnaker}

Deploying the trained SNN on SpiNNaker neuromorphic hardware serves two purposes: it validates that the software-trained spiking dynamics transfer to an asynchronous, event-driven processor, and it enables measurement of true neuromorphic inference energy. The deployment pipeline is shown in Figure~\ref{fig:spinnaker_pipeline}.

\subsection{Hardware Constraints}
\label{ssec:hw-constraints}

