\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}
\label{sec:motivation}

Deep learning has transformed machine perception.
Convolutional neural networks now classify images, transcribe speech, and recognise environmental sounds at or above human-level accuracy on standard benchmarks~\citep{kong2020panns, gong2021ast}.
Yet this capability carries a substantial energy cost.
A single inference through a moderately sized convolutional network on a GPU consumes tens to hundreds of millijoules~\citep{yik2025neurobench}, while the human brain performs real-time, continuous auditory classification within an overall power budget of approximately 20\,W---sustaining on the order of $10^{16}$ synaptic operations per second~\citep{attwell2001energy}.
The disparity between biological and artificial efficiency spans three to four orders of magnitude for equivalent perceptual tasks.

This matters because the most compelling applications for audio intelligence are energy-constrained.
Hearing aids operate on roughly 1\,mW with a weekly charge cycle.
Wildlife monitoring sensors must persist for months on a single battery cell.
Smart building occupancy detectors, industrial fault listeners, and search-and-rescue robots all require continuous audio classification at the edge, where cloud offloading is impractical due to latency, bandwidth, and privacy constraints.
There exists, therefore, a genuine engineering gap: the algorithms that achieve state-of-the-art accuracy are far too expensive to run locally on the devices that need local inference most.

Biology provides an instructive blueprint.
The mammalian auditory system performs real-time, multi-class sound classification at extraordinary energy efficiency through a fundamentally different computational paradigm.
The cochlea decomposes sound into frequency-specific vibrations along the basilar membrane, effectively performing a biological Fourier transform.
Inner hair cells transduce these mechanical vibrations into graded potentials that drive auditory nerve fibres to fire action potentials---binary, all-or-nothing events---at rates and timings that encode frequency, intensity, and temporal structure.
The auditory cortex processes these sparse temporal spike trains through layered circuits of neurons, each of which integrates evidence over time and fires only when its accumulated input exceeds a threshold.
Four properties underpin the efficiency of this system: (i)~binary communication, where neurons transmit one-bit events rather than continuous values; (ii)~temporal sparsity, where only a small fraction of neurons is active at any given moment; (iii)~event-driven processing, where computation occurs only when spikes arrive, not at a fixed clock rate; and (iv)~adaptive thresholding, where neurons accumulate evidence before committing to an output, providing natural noise filtering.

\Glspl{snn} attempt to engineer these properties into artificial systems.
Each neuron maintains a membrane potential that integrates weighted input spikes over time; when this potential crosses a threshold, the neuron emits a spike and resets.
The most widely used neuron model is the \gls{lif} neuron, governed by
\begin{equation}
    \tau_m \frac{dV}{dt} = -V(t) + R \sum_{i} w_i\, s_i(t),
    \label{eq:lif}
\end{equation}
where $V(t)$ is the membrane potential, $\tau_m$ is the membrane time constant, $R$ is the membrane resistance, $w_i$ are synaptic weights, and $s_i(t) \in \{0,1\}$ are binary input spikes.
The energy implication of this formulation is direct.
\Gls{ann} inference requires a \gls{mac} for every active connection: multiply an activation by a weight and accumulate the result.
At 45\,nm technology, a 32-bit floating-point \gls{mac} costs approximately 4.6\,pJ~\citep{horowitz2014energy}.
In an \gls{snn}, binary spikes reduce each operation to a conditional accumulate (\gls{ac}): either add the weight or do nothing.
An \gls{ac} costs approximately 0.9\,pJ---a 5.1$\times$ per-operation advantage~\citep{horowitz2014energy, yik2025neurobench}.
Moreover, temporal sparsity means that the majority of these operations are skipped entirely; \gls{snn} energy scales with the number of spikes emitted, not with the number of parameters.

Dedicated neuromorphic hardware exploits precisely these properties.
SpiNNaker~\citep{furber2014spinnaker}, developed at the University of Manchester, routes binary spike packets between simulated neuron populations using a custom multicast network-on-chip, with each spike event consuming on the order of nanojoules.
Intel's Loihi~2 and IBM's TrueNorth operate on similar event-driven principles.
These platforms offer the potential for orders-of-magnitude energy reduction over equivalent GPU computation, provided that the deployed networks are sufficiently sparse.

For most of the past decade, \glspl{snn} significantly underperformed \glspl{ann} because training was intractable: spike emission is a non-differentiable binary threshold, so standard backpropagation cannot compute gradients through it.
The breakthrough of surrogate gradient descent~\citep{neftci2019surrogate} resolved this impasse.
The key insight is to use exact binary spikes in the forward pass but approximate the spike derivative with a smooth surrogate function (such as an arctangent or a fast sigmoid) in the backward pass.
This permits full backpropagation through time on spiking networks, enabling training on the same datasets and with the same loss functions as conventional networks.
Since 2019, surrogate gradient training has become the dominant approach, producing \glspl{snn} within a few percentage points of matched \glspl{ann} on image classification benchmarks~\citep{eshraghian2023snntorch}.

Environmental sound classification is a natural testbed for \gls{snn} research.
Audio spectrograms are structurally compatible with spiking computation: the temporal axis of a mel spectrogram maps directly to \gls{snn} timesteps, and many environmental sounds---door knocks, glass breaking, dog barks---produce naturally sparse spectrograms that match the regime where \glspl{snn} are most efficient.
Equally importantly, the target applications for environmental sound classification (hearing aids, wildlife sensors, smart buildings) demand exactly the energy efficiency that neuromorphic hardware promises.
ESC-50~\citep{piczak2015esc50}, the standard benchmark for this domain, provides 2{,}000 five-second recordings across 50 classes with predefined five-fold cross-validation, human performance of 81.3\%, and a current state-of-the-art of 99.1\% achieved by OmniVec2 with large-scale pretraining~\citep{srivastava2024omnivec2}.

Despite active research on both \glspl{snn} and audio classification, a verified literature search reveals that \textbf{no prior peer-reviewed work has evaluated convolutional \glspl{snn} on the full 50-class ESC-50 benchmark}.
The closest published result is Larroza et al.~\citep{larroza2025}, who evaluate fully connected (not convolutional) \glspl{snn} on the 10-class ESC-10 subset, achieving approximately 60\% accuracy.
SpiNNaker has been applied to audio classification only with synthetic pure tones~\citep{dominguezmorales2016}.
No systematic comparison of spike encoding methods for audio exists, and no adversarial robustness analysis for audio \glspl{snn} has been published.
This thesis addresses all of these gaps.

\section{Research Questions}
\label{sec:research-questions}

This thesis is structured around four research questions, each targeting a distinct open problem in neuromorphic audio processing.

\begin{description}[style=nextline,leftmargin=1.5cm,labelwidth=1cm]

\item[\textbf{RQ1}] \textit{How does \gls{snn} accuracy compare to a matched \gls{ann} on ESC-50?}

The relevant comparison is not against state-of-the-art systems pretrained on millions of external clips, but against an \gls{ann} with identical architecture, training protocol, and data budget.
The measured accuracy gap under these controlled conditions quantifies the genuine cost of spiking computation.
This thesis finds that the gap is 16.70 percentage points from scratch (47.15\% versus 63.85\%, $p = 0.001$), but collapses to 0.95 percentage points when both architectures classify pretrained audio features via \gls{panns} (92.50\% versus 93.45\%, $p = 0.034$).

\item[\textbf{RQ2}] \textit{Which spike encoding method performs best for environmental sound classification, and why?}

Static mel spectrograms must be converted into temporal spike trains for \gls{snn} processing.
Seven encoding methods---rate, latency, delta, direct, burst, phase, and population---are evaluated under identical conditions.
The goal is not merely to rank them, but to explain the ranking mechanistically in terms of information preservation, temporal coverage, and loss landscape geometry.
The thesis establishes the hierarchy: direct (47.15\%) $\gg$ rate (24.00\%) $\approx$ phase (24.15\%) $>$ population (19.15\%) $>$ latency (16.30\%) $\gg$ delta (7.25\%) $\approx$ burst (6.50\%).

\item[\textbf{RQ3}] \textit{Can a trained \gls{snn} be deployed on SpiNNaker neuromorphic hardware, and what is the accuracy cost?}

SpiNNaker imposes hard constraints---binary spike communication, fixed-point arithmetic, no native convolutional support---that create significant engineering challenges for deploying a float-trained network.
This thesis documents the complete deployment pipeline, including the failures, their root causes, and the hybrid current-injection approach that achieves 57.35\% ($\pm$2.39\%) on hardware with a gap of only 2.15 percentage points relative to software simulation.

\item[\textbf{RQ4}] \textit{What advantages do \glspl{snn} exhibit beyond raw classification accuracy?}

Prior work on image-domain \glspl{snn} suggests that binary thresholding may confer natural adversarial robustness~\citep{sharmin2020inherent}.
This thesis extends that investigation to audio spectrograms, finding 6.0$\times$ greater robustness under FGSM attack ($\varepsilon = 0.1$: \gls{snn} 16.55\% versus \gls{ann} 2.75\%).
Additionally, continual learning experiments reveal that the \gls{snn} suffers 4.7 percentage points less catastrophic forgetting than the matched \gls{ann}, and a pruning sweep demonstrates that 85\% weight pruning maintains \gls{snn} accuracy while reducing energy to 27\,nJ per inference---16.6$\times$ cheaper than the \gls{ann}.

\end{description}

\section{Contributions}
\label{sec:contributions}

This thesis makes six original contributions:

\begin{enumerate}[label=\textbf{C\arabic*},leftmargin=2cm]

\item \textbf{First convolutional \gls{snn} benchmark on ESC-50.}
A SpikingCNN (two convolutional blocks, two fully connected layers, ${\sim}$622K parameters) achieves 47.15\% ($\pm$4.50\%) on the full 50-class ESC-50 benchmark with five-fold cross-validation using direct encoding.
The 16.70 percentage point gap relative to a matched \gls{ann} (63.85\% $\pm$ 3.07\%) is characterised mechanistically and contextualised through the \gls{panns} transfer experiment.

\item \textbf{Most comprehensive spike encoding comparison for audio.}
Seven encoding methods are evaluated under identical architecture, training, and evaluation conditions.
The resulting hierarchy is explained in terms of information preservation: direct encoding dominates because it feeds continuous magnitudes at every timestep, while burst and delta coding fail catastrophically due to temporal window mismatch and static-spectrogram incompatibility, respectively.
The finding that phase coding (one deterministic spike per neuron) matches rate coding (approximately seven stochastic spikes per neuron) at identical accuracy has direct implications for energy-efficient deployment.

\item \textbf{Rhythm-SNN: oscillatory threshold modulation for improved accuracy.}
A novel technique that modulates \gls{lif} firing thresholds with a learnable sinusoidal signal improves scratch-trained \gls{snn} accuracy from 47.15\% to 61.10\% ($\pm$1.99\%), reducing the \gls{snn}--\gls{ann} gap from 16.70 to 2.75 percentage points.
This represents the highest reported convolutional \gls{snn} accuracy on ESC-50.

\item \textbf{SpiNNaker deployment for environmental sound classification.}
The Rhythm-SNN is deployed on SpiNNaker~2 neuromorphic hardware via a current-injection hybrid approach, achieving 57.35\% ($\pm$2.39\%) across five folds with only a 2.15 percentage point hardware--software gap.
A pruning sweep across 50 hardware deployments (10 sparsity levels $\times$ 5 folds) maps the accuracy--energy Pareto frontier on real neuromorphic silicon.

\item \textbf{Pruning--energy Pareto analysis on neuromorphic hardware.}
Using NeuroBench-compliant energy estimation~\citep{yik2025neurobench}, a systematic pruning sweep from 50\% to 95\% sparsity demonstrates that 85\% pruning maintains classification accuracy while reducing estimated inference energy to 27\,nJ per sample---16.6$\times$ cheaper than the equivalent \gls{ann} at 454\,nJ.

\item \textbf{Adversarial robustness and continual learning advantages.}
The \gls{snn} demonstrates 6.0$\times$ greater adversarial robustness under \gls{fgsm} attack at $\varepsilon = 0.1$ (16.55\% versus 2.75\%) and 195$\times$ under \gls{pgd} at $\varepsilon = 0.05$ (9.75\% versus 0.05\%), validated across five folds.
In sequential continual learning, the \gls{snn} exhibits 4.7 percentage points less catastrophic forgetting than the matched \gls{ann} (69.9\% versus 74.7\%).

\end{enumerate}

\section{Evaluation Strategy}
\label{sec:evaluation-strategy}

The experimental programme follows two established evaluation frameworks to ensure rigour and reproducibility.

\paragraph{Classification accuracy.}
All accuracy results use ESC-50's predefined five-fold cross-validation protocol~\citep{piczak2015esc50}, wherein four folds serve as training data and the fifth as a held-out test set, rotating across all five partitions.
This protocol is mandatory for comparability with published ESC-50 results.
Reported accuracies are the best validation accuracy per fold (not the final epoch), and all means and standard deviations are computed across the five folds.
Statistical significance is assessed using paired $t$-tests across folds, with $p < 0.05$ as the threshold.

\paragraph{Energy estimation.}
Energy analysis follows the NeuroBench framework~\citep{yik2025neurobench}, which defines standardised metrics for neuromorphic system comparison.
The key metrics are \textit{Effective \glspl{ac}} for \glspl{snn} (counting only non-zero spike-weight accumulations) and \textit{Effective \glspl{mac}} for \glspl{ann} (counting only non-zero activation-weight multiplications).
Energy is estimated using the Horowitz cost model~\citep{horowitz2014energy}: 0.9\,pJ per \gls{ac} and 4.6\,pJ per \gls{mac} at 45\,nm.
All energy figures are five-fold validated using NeuroBench~v2.2.0.
While these are software-simulation proxies rather than direct hardware power measurements, the NeuroBench protocol enables standardised comparison across systems and is the accepted community standard.

\paragraph{Adversarial robustness.}
Robustness is evaluated using \gls{fgsm}~\citep{goodfellow2015fgsm} and \gls{pgd}~\citep{madry2018pgd} attacks at seven perturbation magnitudes ($\varepsilon \in \{0.001, 0.005, 0.01, 0.02, 0.05, 0.1, 0.2\}$), applied identically to the \gls{snn} and matched \gls{ann}.
The robustness ratio---retained \gls{snn} accuracy divided by retained \gls{ann} accuracy at each $\varepsilon$---provides a scale-invariant comparison.

\paragraph{Continual learning.}
Catastrophic forgetting is measured following the sequential task protocol: ESC-50's five super-categories (animals, nature, human, domestic, urban) are presented one at a time, and backward transfer is computed as the mean accuracy drop on previously learned tasks after training on subsequent ones.

\section{Report Structure}
\label{sec:report-structure}

The remainder of this report is organised as follows.

\textbf{Chapter~\ref{ch:background}} reviews the relevant background and related work, covering ESC-50 and environmental sound classification, \gls{lif} neuron dynamics and surrogate gradient training, the seven spike encoding methods with their mathematical formulations, prior \gls{snn} work on audio, the SpiNNaker architecture, energy efficiency frameworks, adversarial robustness in \glspl{snn}, \gls{panns} transfer learning, and continual learning.

\textbf{Chapter~\ref{ch:methodology}} details the experimental design philosophy and controlled comparison framework, the dataset and preprocessing pipeline, both network architectures (SpikingCNN and ConvANN), all seven encoding implementations, the training protocol, the SpiNNaker deployment strategy, and the design of advanced experiments.

\textbf{Chapter~\ref{ch:results}} presents all experimental results: the \gls{ann} baseline, the seven-encoding comparison, the Rhythm-SNN and advanced techniques, SpiNNaker hardware deployment with pruning sweep, energy analysis, \gls{panns} transfer learning, adversarial robustness, continual learning, and the surrogate gradient ablation.

\textbf{Chapter~\ref{ch:discussion}} synthesises the findings, explains the encoding hierarchy through an information preservation principle, discusses SpiNNaker co-design insights, analyses the adversarial robustness mechanism, and acknowledges threats to validity.

\textbf{Chapter~\ref{ch:conclusion}} summarises all six contributions, provides concise answers to the four research questions, and identifies directions for future work.
